---
title: "BioDataAnalysis"
author: "Axel Muniz Tello"
date: "2025-11-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load the Dataset
dataframe <- read.csv("/Users/axeltello/Desktop/BioData.csv") # change directory of file to your local file
head(dataframe)
```

```{r}
summary(dataframe[, 1:7]) # samples of morgan predictors [0,5]
```

# Regression Model (1) -- Standard Linear Regression :
```{r}
standard_regression <- lm(pIC50 ~ ., data = dataframe[, -1]) # linear regression model
```

Show the summary statistics of the regression model
```{r}
summary(standard_regression)
```

```{r}
# Plot performance for standard regression
plot(standard_regression)

# Comments: 
# ---------
# Multiple points on the line indicate colinearity in the data due to the sparsity with limited excitation
# need a model that will be treated as a regression model, but classified using a log function for determining 
# factors of valley fever.
```
### Computation Notes: 
- Most rows have a NA or NaN $\implies$ multicollinearity, and no prediction
- We will use LASSO instead to see how it can handle the dataset

# Regression Model (2) -- Lasson Regression Model:
```{r}
library(glmnet)

## 1) start from your original data
df <- dataframe

## 2) make sure pIC50 is numeric
df$pIC50 <- as.numeric(df$pIC50)

## 3) drop rows with missing target
df <- df[!is.na(df$pIC50), ]

## 4) grab only the morgan_* predictors
x_all <- df[, grepl("^morgan_", names(df)), drop = FALSE]

## 5) drop predictors that are constant (all 0 or all 1)
varying <- apply(x_all, 2, function(z) length(unique(z)) > 1)
x <- as.matrix(x_all[, varying, drop = FALSE])

## 6) response
y <- df$pIC50

## 7) fit LASSO with CV
set.seed(1)
cv_model <- cv.glmnet(x, y, alpha = 1)     # LASSO
best_lambda <- cv_model$lambda.min

## 8) final model at best lambda
lasso <- glmnet(x, y, alpha = 1, lambda = best_lambda)

## 9) look at coefficients (nonzero ones are the important bits)
coef(lasso, s = 0.0001)


```

```{r}
plot(lasso)

# Computational Notes: 
# --------------------
# Not enough data is able to be modeled because we run into the same issue with sparsity of data and low excitation
```

```{r}
summary(lasso) # useless information
```


 
# Regression Model (2) -- Ridge Regression Model:
```{r}
library(glmnet)

X <- as.matrix(df[, grepl("^morgan_", names(df))])
y <- df$pIC50

set.seed(1)
cv_ridge <- cv.glmnet(X, y, alpha = 0)   # alpha=0 → Ridge
best_lambda_ridge <- cv_ridge$lambda.min

ridge_model <- glmnet(X, y, alpha = 0, lambda = best_lambda_ridge)
coef(ridge_model)

```
```{r}
plot(ridge_model)
```

# Regression Model (3) -- Elastic: 
```{r}
library(glmnet)

X <- as.matrix(df[, grepl("^morgan_", names(df))])
y <- df$pIC50  # or 0/1 if classification

cv_model <- cv.glmnet(X, y, alpha = 0.5)  # mix of LASSO + Ridge
best_lambda <- cv_model$lambda.min
model <- glmnet(X, y, alpha = 0.5, lambda = best_lambda)

coef_table <- as.matrix(coef(model))
nonzero <- coef_table[coef_table[,1] != 0, , drop=FALSE]
head(nonzero)

plot(cv_model)

# Computational Notes: 
# --------------------
# Console provides intuitive possible markers for fighting valley fever. 
# elastic balances both ridge and and lassso with alpha set to 0.5
```

```{r}
summary(cv_model)
```


